Advancing spatially explicit fisheries management with age-structured species distribution models

### 2025-01-28 paper.qmd
editing text

Species distribution models (SDMs) are widely used in fisheries and ecology, providing insights into species' spatial distributions. While many SDMs incorporate external information, such as environmental or habitat features, the majority overlook age-specific information, despite the known influence of age structure on fish distribution and abundance. Including age information may be particularly valuable for species exhibiting age-related spatial patterns driven by ontogenetic shifts, recruitment dynamics, and selective fishing pressures. Here, we develop an age-structured SDM framework using data from the west coast of the USA, and concentrate on two species with distinct life histories: North Pacific hake (*Merluccius productus*) and sablefish (*Anoplopoma fimbria*). Our framework uses spatiotemporal models to predict age-specific distributions. We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlight that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes. Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2010, 2014 and 2016 cohorts; sablefish from 2008, 2010, 2016, and 2021 cohorts). Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk in space, helping increase the efficiency of fisheries. Our framework is applicable to a wide range of survey types and platforms around the world, and supports the development of spatially explicit management strategies and optimal allocation of fishing effort.

@@

Incorporating age structure into SDMs may provide more accurate predictions by capturing spatial and temporal variability related to life history traits, age-specific habitat preferences, and population dynamics. Fish often exhibit ontogenetic shifts, where younger individuals occupy different habitats than adults due to factors like predator avoidance, habitat preferences, and diet changes [@werner1984; @ciannelli2004]. These shifts can create age-specific distributions across habitats, particularly for species that undergo distinct life history phases. Additionally, recruitment pulses and migration patterns vary by age, influencing the spatial distribution of younger versus older fish. Fishing selectivity often further accentuates these patterns by disproportionately removing larger, older fish, which alters the population structure and potentially shifts spatial distributions [@hilborn1992; @methot2013]. By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relati

@@

On the west coast of the USA, the West Coast Groundfish Bottom Trawl Survey (WCGBTS) has been an annual survey conduced from 2003 - present. The WCGBTS is designed to estimate the abundance, size, and age composition of groundfish species important to commercial and recreational fisheries found in near-bottom habitats on the west coast of the USA (Keller et al., 2017). The survey effort is concentrated in summer months and has been conducted annually since 2003 (here we use the data through 2018; data are publicly available at https://www.nwfsc.noaa.gov/data). Importantly, the random stratified sampling design, spatial and seasonal coverage, effort, and gears have remained relatively constant within the period we analyze. Like many other surveys around the world, no WCGBTS survey occurred in 2020 because of the Covid-19 pandemic. Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish. \[more on biological gradients\] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years. Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada (Grandin et al. 2024) and sablefish is frequently assessed because of its high commercial value. For our analysis, we used hake collected 2007 -- 2019 (mean 650.7 individuals sampled per year) and sablefish collected 2003 -- 2023 (mean 1314.4 individuals sampled per year).

@@

```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent strong cohorts in our dataset (2010, 2014, 2016). CPUE is standardized to 1.0 across cohorts for visualization purposes. Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
knitr::include_graphics("plots/Pacific hake-follow-age-class.png")
```

>>>>

Species distribution models (SDMs) are widely used in fisheries and ecology, providing insights into species' spatial distributions. While many SDMs incorporate external information, such as environmental or habitat features, the majority overlook age-specific information, despite the known influence of age structure on fish distribution and abundance. Including age information may be particularly valuable for species exhibiting age-related spatial patterns driven by ontogenetic shifts, recruitment dynamics, and selective fishing pressures. Here, we develop an age-structured SDM framework using data from the west coast of the USA, and concentrate on two species with distinct life histories: North Pacific hake (*Merluccius productus*) and sablefish (*Anoplopoma fimbria*). Our framework uses spatiotemporal models to predict age-specific distributions. We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlight that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes. Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2008, 2010, 2014 and 2016 cohorts; sablefish from 2008, 2010, 2016, and 2021 cohorts). Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk in space, helping increase the efficiency of fisheries. Our framework is applicable to a wide range of survey types and platforms around the world, and supports the development of spatially explicit management strategies and optimal allocation of fishing effort.

@@

Incorporating age structure into SDMs may provide more accurate predictions by capturing spatial and temporal variability related to life history traits, age-specific habitat preferences, and population dynamics. Fish often exhibit ontogenetic shifts, where younger individuals occupy different habitats than adults due to factors like predator avoidance, habitat preferences, and diet changes [@werner1984; @ciannelli2004]. These shifts can create age-specific distributions across habitats, particularly for species that undergo distinct life history phases. Additionally, spatiallly variable recruitment pulses and age-related migration patterns can influence the spatial distribution of younger versus older fish. Fishing selectivity often further accentuates these patterns by disproportionately removing larger, older fish, which alters the population structure and potentially shifts spatial distributions [@hilborn1992; @methot2013]. By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relationships.

@@

On the west coast of the USA, the West Coast Groundfish Bottom Trawl Survey (WCGBTS) has been an annual survey conduced from 2003 - present. The WCGBTS is designed to estimate the abundance, size, and age composition of groundfish species important to commercial and recreational fisheries found in near-bottom habitats on the west coast of the USA [@keller2017]. The survey effort is concentrated in summer months and has been conducted annually since 2003 (here we use the data through 2018; data are publicly available at https://www.nwfsc.noaa.gov/data). Importantly, the random stratified sampling design, spatial and seasonal coverage, effort, and gears have remained relatively constant within the period we analyze. Like many other surveys around the world, no WCGBTS survey occurred in 2020 because of the Covid-19 pandemic. Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish. [more on biological gradients] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years. Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada [@Grandin2024] and sablefish is frequently assessed because of its high commercial value. For our analysis, we used hake collected 2007 -- 2019 (mean 650.7 individuals sampled per year) and sablefish collected 2003 -- 2023 (mean 1314.4 individuals sampled per year).

@@

```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent strong cohorts in our dataset (2008, 2010, 2014, 2016). CPUE is standardized to 1.0 across cohorts for visualization purposes. Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
knitr::include_graphics("plots/Pacific hake-follow-age-class.png")
```
